Approximating the end of nested sampling

### 2023-07-27 paper/paper.tex
added anatomy figure to paper

\title[Approximating the end of nested sampling]{Approximating the end of nested sampling}
\author[Z. Hu]{Zixiao Hu, Will Handley}

@@

\begin{abstract}
This project develops a technique to estimate the runtime of a nested sampling run at an intermediate iteration. The known likelihood at an iteration is fitted, then extrapolated using a Gaussian model likelihood, which allows the remaining evidence and final prior volume to be calculated. We find that the method succesfully recovers the true endpoint within standard error from the beginning for near-Gaussian likelihoods, and by the halfway point for non-Gaussian likelihoods while identifying the correct order of magnitude throughout. 
\end{abstract}

@@

\subsubsection{Endpoint formula}
Now we are equipped with the tools to make endpoint predictions. We will begin by fitting the model over the live points only. At each iteration, estimate the prior volumes of the live points to be those which correspond to killing off each point one by one without replacement and using formula \eqref{X_distribution}. Combined with the live likelihoods $\logL$, the parameters of the model can be found by minimising the cost function
\begin{equation}\label{chi squared}
	C^2(\thetab) = \sum_i \left| \log \Like_i - \log f(X_i, \thetab) \right| ^2
\end{equation}
with respect to $\thetab$. A key feature is that we can analytically differentiate  $C^2$ for the Gaussian model, which allows us to write the optimal values of $\logLm$ and $\sigma$ in terms of the optimal values of $d$:
\begin{equation}
    \sigma^2 = \frac{N \sum_i X_i^{4/d} - \left(\sum_i X_i^{2/d}\right)^2}{2 \sum_i \log \Like_i \sum_i X_i^{2/d} - 2N \sum_i \log \Like_i X_i^{2/d}}
\end{equation}
\begin{equation}
    \logLm = \frac{1}{N} \sum_i \logL + \frac{1}{2N\sigma^2} \sum_i X_i^{2/d} 
\end{equation}
(where $N$ here is the number of data points in we fit over), and simply minimise with respect to $d$ instead of all three variables. The result is that we are able to consistently find \textit{global} rather than local minima in a single optimisation run. We do not need to concern ourselves with the usual difficulties of global optimisation like comparing an ensemble of initial conditions, which is orders of magnitude slower.
\par
Once we have obtained the best-fit $\thetab$, the endpoint can be calculated in the manner described in \eqref{model endpoint};
\begin{equation}
	\epsilon = \frac{\int_0^{X_f} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ dX}{\int_0^{X_i} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ dX + Z_\mathrm{dead}}.
\end{equation}
The integrals have the analytic solution
\begin{equation}
	\int_0^{X_k} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ dX = \frac{d}{2} \cdot 2^{d/2} \cdot \sigma^{d} \cdot \gamma_k
\end{equation}
where $\gamma_k = \gamma\left(d/2, X_k^{2/d}/2\sigma^2\right)$ is the lower incomplete gamma function. After taking the inverse of  $\gamma$ and a few more steps of algebra, we arrive at
\begin{multline}
	\log X_f = \frac{d}{2}\log 2 + d \log \sigma \\
	+ \log \gamma^{-1} \left( d/2 ,\ \epsilon \left(\gamma_i+ 2^{-d/2} \sigma^{-d} Z_\mathrm{dead}/\Like_\mathrm{max}\right)\right),
\end{multline}
and $N_f$ is just  $-\nlive$ multiplied by this. 
\par
An useful feature is that the formula is robust when we are far away from the posterior bulk. In this phase, the inferred values of $\Like_\mathrm{max}$ turn out to be very uncertain and often divergent, but the dead evidence is also negligible, so $\Like_\mathrm{max}$ has little impact on the prediction. Once $Z_\mathrm{dead}$ becomes comparable to the live evidence $\Like_\mathrm{max}$ does begin to impact the prediction, but at this point we have reached the posterior bulk, so we expect $\Like_\mathrm{max}$ to be well-estimated and thus contribute in a useful way.
\par
Note that the prediction procedure should not contribute any significant time complexity to the nested sampling run, since it involves no likelihood evaluations.

>>>>

\title[Approximating the end of nested sampling]{Approximating the end of nested sampling}
\author[Z. Hu et. al]{Zixiao Hu, Artyom Baryshnikov, Will Handley}

@@

\begin{abstract}
This paper develops a technique to estimate the runtime of a nested sampling run at an intermediate iteration for any nested sampler. The known likelihood at an iteration is fitted, then extrapolated using a Gaussian model likelihood, which allows the remaining evidence and prior volume at termination to be calculated. We find that the method succesfully recovers the true endpoint within standard error from the beginning for near-Gaussian likelihoods. For non-Gaussian likelihoods, the estimate is still the correct order of magnitude and converges to the correct answer by the end of the run. 
\end{abstract}

@@

\subsubsection{Endpoint formula}
Now we are equipped with the tools to make endpoint predictions. We will begin by fitting the model over the live points only. At each iteration, estimate the prior volumes of the live points to be those which correspond to killing off each point one by one without replacement and using formula \eqref{X_distribution}. Combined with the live likelihoods $\logL$, the parameters of the model can be found by minimising the cost function
\begin{equation}\label{chi squared}
	C^2(\thetab) = \sum_i \left| \log \Like_i - \log f(X_i, \thetab) \right| ^2
\end{equation}
with respect to $\thetab$. A key feature is that we can analytically differentiate  $C^2$ for the Gaussian model, which allows us to write the optimal values of $\logLm$ and $\sigma$ in terms of the optimal values of $d$:
\begin{equation}
    \sigma^2 = \frac{N \sum_i X_i^{4/d} - \left(\sum_i X_i^{2/d}\right)^2}{2 \sum_i \log \Like_i \sum_i X_i^{2/d} - 2N \sum_i \log \Like_i X_i^{2/d}}
\end{equation}
\begin{equation}
    \logLm = \frac{1}{N} \sum_i \logL + \frac{1}{2N\sigma^2} \sum_i X_i^{2/d} 
\end{equation}
(where $N$ here is the number of data points in we fit over), and simply minimise with respect to $d$ instead of all three variables. The result is that we are able to consistently find \textit{global} rather than local minima in a single optimisation run. We do not need to concern ourselves with the usual difficulties of global optimisation like comparing an ensemble of initial conditions, which is orders of magnitude slower.
\par
Once we have obtained the best-fit $\thetab$, the endpoint can be calculated in the manner described in \eqref{model endpoint};
\begin{equation}
	\epsilon = \frac{\int_0^{X_f} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ dX}{\int_0^{X_i} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ dX + Z_\mathrm{dead}}.
\end{equation}
The integrals have the analytic solution
\begin{equation}
	\int_0^{X_k} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ dX = \frac{d}{2} \cdot 2^{d/2} \cdot \sigma^{d} \cdot \gamma_k
\end{equation}
where $\gamma_k = \gamma\left(d/2, X_k^{2/d}/2\sigma^2\right)$ is the lower incomplete gamma function. After taking the inverse of  $\gamma$ and a few more steps of algebra, we arrive at
\begin{multline}
	\log X_f = \frac{d}{2}\log 2 + d \log \sigma \\
	+ \log \gamma^{-1} \left( d/2 ,\ \epsilon \left(\gamma_i+ 2^{-d/2} \sigma^{-d} Z_\mathrm{dead}/\Like_\mathrm{max}\right)\right),
\end{multline}
and $N_f$ is just  $-\nlive$ multiplied by this. 
\par
Noting that the prediction depends on the \textit{ratio} $Z_\mathrm{dead}/\Like_\mathrm{max}$, an useful feature is that the formula is robust when we are far away from the posterior bulk. In this phase, the inferred values of $\Like_\mathrm{max}$ turn out to be very uncertain and often divergent, but the dead evidence is also negligible, so $\Like_\mathrm{max}$ has little impact on the prediction. Once $Z_\mathrm{dead}$ becomes comparable to the live evidence $\Like_\mathrm{max}$ does begin to impact the prediction, but at this point we have reached the posterior bulk, so we expect $\Like_\mathrm{max}$ to be well-estimated and thus contribute in a useful way.
\par
Note that the prediction procedure should not contribute any significant time complexity to the nested sampling run, since it involves no likelihood evaluations.
